---
title: "gsbe-data-prep"
author: "bbl"
format:
  html:
    code-fold: true
toc: true
editor: visual
---

```{r setup}
#| include: false

library(dplyr)
library(tidyr)
library(ggplot2)
theme_set(theme_bw())
library(sf)
library(rnaturalearth)
library(readr)
library(terra)
library(lubridate)
library(DT)
library(arrow)
```

## SRDB

```{r read-srdb}
#| fig-width: 12

read_csv("data-srdb/srdb-data.csv", show_col_types = FALSE) %>% 
    select(Author, Study_midyear, YearsOfData, 
           Latitude, Longitude, Elevation, Manipulation, 
           Biome, Ecosystem_type, Ecosystem_state, 
           Rs_annual, Rh_annual, Rs_growingseason) %>% 
    filter(!is.na(Longitude), !is.na(Latitude), !is.na(Study_midyear)) %>% 
    filter(!is.na(Rs_growingseason) | !is.na(Rs_annual) | !is.na(Rh_annual)) %>% 
    filter(Manipulation == "None") %>% 
    select(-Elevation, -Manipulation) %>% 
    mutate(Longitude = round(Longitude, 3),
           Latitude = round(Latitude, 3)) ->
    srdb

message(nrow(srdb), " rows of data")

message("There are ", sum(srdb$YearsOfData > 1, na.rm = TRUE), " multi-year observations to remove")

srdb %>% 
    filter(YearsOfData == 1) %>% 
    select(-YearsOfData) ->
    srdb
message(nrow(srdb), " final rows of data")

world <- ne_coastline()
srdb_annual <- srdb %>% filter(is.na(Rs_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_annual N=", nrow(srdb_annual)))

srdb_rh_annual <- srdb %>% filter(!is.na(Rh_annual))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_rh_annual, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rh_annual N=", nrow(srdb_rh_annual)))

srdb_growingseason <- srdb %>% filter(is.na(Rs_growingseason))
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_growingseason, 
               aes(x = Longitude, y = Latitude, color = Biome),
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB Rs_growingseason N=", nrow(srdb_growingseason)))
```

```{r srdb-prep}
#| fig-width: 12

# Subsample SRDB data for quick testing
# set.seed(12)
# srdb %>% 
#     slice_sample(n = 20) -> 
#     srdb_sample
srdb_sample <- srdb

srdb_sample %>% 
    select(Longitude, Latitude) ->
    coords

ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = Biome),
               size = 4,
               na.rm = TRUE) + 
    ggtitle(paste0("SRDB subsample N=", nrow(srdb_sample)))
```

## SPEI

We are using the ERA5–Drought dataset; see Keune et al. 2025 <https://www.nature.com/articles/s41597-025-04896-y>.

```{r read-spei}
#| fig-height: 6
#| fig-width: 10

SPEI_CACHE <- "caches/SPEI_data_cache.parquet"
if(file.exists(SPEI_CACHE)) {
    message("Reading data from SPEI cache...")
    spei <- arrow::read_parquet(SPEI_CACHE)    
} else {
    
    process_spei <- function(which_spei, coords) {
        files <- list.files(path = "data-spei/", 
                            pattern = paste0("^", which_spei, "_.*nc$"), 
                            full.names = TRUE)
        message("Processing ", length(files), " ", which_spei, " files...")
        spei_list <- list()
        for(f in files) {
            x <- suppressWarnings(terra::rast(f)) 
            dat <- terra::extract(x, coords)
            dat$time <- terra::time(x)
            spei_list[[f]] <- dat
        }
        bind_rows(spei_list) %>% 
            as_tibble() %>% 
            return()
    }
    
    # The SPEI index comes in different 'accumulation periods' of drought;
    # we're interested in 12, 24, 36, and 48 months
    spei12 <- process_spei("SPEI12", coords)
    #spei24 <- process_spei("SPEI24", coords)
    spei <- spei12
    spei$Biome <- srdb_sample$Biome[spei$ID]
    
    arrow::write_parquet(spei, SPEI_CACHE)
}

# Distribution plots
ggplot(spei, aes(x = SPEI12)) +
    geom_histogram(bins = 50, na.rm = TRUE) + 
    facet_wrap(~Biome) +
    ggtitle("SPEI12")

ggplot(spei, aes(time, SPEI12, color = Biome, group = ID)) + 
    geom_line(na.rm = TRUE) + 
    facet_grid(Biome~.) + 
    theme(legend.position = "none") +
    ggtitle("SPEI12")

# Summarise to annual, January-March, and July-September
spei %>% 
    mutate(Year = year(time), 
           Month = month(time)) %>% 
    arrange(time) %>% 
    group_by(Biome, ID, Year) %>% 
    summarise(SPEI12jja = round(mean(SPEI12[7:9]), 3),
              SPEI12jfm = round(mean(SPEI12[1:3]), 3),
              SPEI12 = round(mean(SPEI12), 3),
              .groups = "drop") ->
    spei_annual

# Summary plots
ggplot(spei_annual, aes(Year, SPEI12, color = Biome, group = ID)) + 
    geom_line(na.rm = TRUE) + 
    facet_grid(Biome~.) + 
    theme(legend.position = "none") +
    ggtitle("SPEI12 - annual mean")

ggplot(spei_annual, aes(Year, SPEI12jja, color = Biome, group = ID)) + 
    geom_line(na.rm = TRUE) + 
    facet_grid(Biome~.) + 
    theme(legend.position = "none") +
    ggtitle("SPEI12 - jja")

message("Merging into SRDB data...")
srdb_sample$ID <- seq_len(nrow(srdb_sample))
srdb_sample$Year <- floor(srdb_sample$Study_midyear)
srdb_sample$Year1 <- srdb_sample$Year - 1

spei_annual %>% select(ID, Year, SPEI12) -> spei_annual_join
spei_annual_join %>% rename(SPEI12_y1 = SPEI12) -> spei_annual_join_y1

srdb_sample %>% 
    left_join(spei_annual_join, by = c("ID" = "ID", "Year" = "Year")) %>% 
    left_join(spei_annual_join_y1, by = c("ID" = "ID", "Year1" = "Year")) ->
    srdb_sample

ggplot(srdb_sample, aes(x = SPEI12_y1)) + 
    geom_histogram(bins = 50, na.rm = TRUE) + 
    geom_vline(xintercept = -0.84, linetype = 2, color = "red") +
    geom_vline(xintercept = -1.28, linetype = 2, color = "red") +
    geom_vline(xintercept = -1.65, linetype = 2, color = "red") +
    facet_wrap(~Biome) +
    ggtitle("SPEI12 previous year", 
            subtitle = "Dashed lines show SPEI categories for mild, moderate, and severe dryness")

```

```{r spei-over-time}
#| fig-height: 14
#| fig-width: 10

ggplot(srdb_sample, aes(Year, SPEI12_y1, color = Biome)) + 
    geom_hline(yintercept = -0.84, linetype = 2, color = "red") +
    geom_hline(yintercept = -1.28, linetype = 2, color = "red") +
    geom_hline(yintercept = -1.65, linetype = 2, color = "red") +
    geom_point(na.rm = TRUE) + 
    facet_grid(Biome~.) + 
    theme(legend.position = "none") +
    ggtitle("SPEI12 previous year",
            subtitle = "Dashed lines show SPEI categories for mild, moderate, and severe dryness")

# Are we identifying SPEI jumps in SRDB correctly?
srdb_sample %>% 
    filter(SPEI12 > 0 & SPEI12_y1 <= -1) %>% 
    slice_sample(n = 10) -> 
    srdb_pbe # "potential birch effect"

spei_annual %>% 
    filter(ID %in% srdb_pbe$ID) %>% 
    left_join(select(srdb_pbe, ID, SRDB_Year = Year), by = "ID") %>% 
    ggplot(aes(Year, SPEI12, color = SRDB_Year == Year)) + 
    geom_point() + 
    facet_grid(ID ~ .) +
    ggtitle("Are we identifying SPEI jumps in SRDB correctly? (random 10)")

```

## Climate data

Ugh, the ERA5 data are *big* and operations are *slow*. I'm not a GIS expert, so spent a while playing around with `terra::crop()`, `raster::aggregate()`, etc., with little success.

We use two strategies to deal with this:

-   Download individual-year `grib` files from the EMCWF Climate Data Store (a pain to download, but then extracting from them is relatively fast); and

-   Use a cache so we only process things once, hopefully

```{r read-era5}
#| fig-height: 6
#| fig-width: 10

# Load cached data if available
ERA5_SHORT_NAME <- read_csv("ERA5_short_names.csv", col_types = "cc")
ERA5_CACHE <- "caches/ERA5_data_cache.parquet"
if(file.exists(ERA5_CACHE)) {
    message("Reading cache file ", ERA5_CACHE)
    cache <- arrow::read_parquet(ERA5_CACHE)    
} else {
    message("No ERA5 data cache found")
    cache <- tibble(Year = numeric(0), Longitude = numeric(0), Latitude = numeric(0))
}

# Get distinct year/location combinations -- that's we want to pull out of ERA5 files
srdb_sample %>% 
    distinct(Year, Longitude, Latitude) ->
    srdb_yrlonlat

cache_hits <- 0
for(i in seq_len(nrow(srdb_yrlonlat))) {
    yr <- srdb_yrlonlat$Year[i]
    lon <- srdb_yrlonlat$Longitude[i]
    lat <- srdb_yrlonlat$Latitude[i]
    
    # Are data already in cache?
    cache %>% 
        filter(Year == yr, Longitude == lon, Latitude == lat) ->
        dat
    if(nrow(dat) > 0) {
        cache_hits <- cache_hits + 1
        #message("\t", nrow(dat), " data rows are in cache")
    } else {
        if(cache_hits > 0) { 
            message("(", cache_hits, " cache hits)")
            cache_hits <- 0
        }
        message(i, "/", nrow(srdb_yrlonlat), ": ", yr, " ", lon, " ", lat, appendLF = FALSE)
        # Construct year-specific file name; these were downloaded from ECMWF
        # and contain 2m temperature, soil temperature levels 1 and 2, 
        # volumetric soil water content levels 1 and 2, total precipitation,
        # and LAI of high vegetation
        f <- file.path("data-ERA5/", paste0("ERA5_", yr, ".grib"))
        if(!file.exists(f)) {
            message("\t", f, " does not exist; skipping")
            next
        }
        message("\tNo data in cache; extracting from ", f, "...")
        x <- terra::rast(f)
        
        # Make sure the time period is what we're expecting
        if(!all(year(time(x)) == yr)) {
            stop("Year is supposed to be ", yr, " but file data dates are different!")
        }
        
        # Extract, reshape, clean up
        y <- terra::extract(x, srdb_yrlonlat[i,][-1]) # this is the expensive step
        y %>% 
            pivot_longer(-ID) %>% 
            mutate(time = time(x), Year = yr, Longitude = lon, Latitude = lat) %>% 
            left_join(ERA5_SHORT_NAME, by = "name") %>% 
            select(Year, time, Longitude, Latitude, short_name, value) ->
            dat
        cache <- bind_rows(cache, dat)
        message("\tAdding ", nrow(dat), " data rows to cache ", ERA5_CACHE, "...")
        arrow::write_parquet(cache, ERA5_CACHE)
    }
}

# At this point the cache data frame has all the needed data
# Reshape, summarise to annual (NB this may change in the future),
# and join with SRDB data
message("Summarizing and joining with SRDB data...")
cache %>% 
    mutate(time = as.Date(time)) %>% 
    pivot_wider(names_from = "short_name") %>% 
    group_by(Year, Longitude, Latitude) %>%
    summarise(Tair = round(mean(Tair) - 273.1, 2),
              Tsoil_lev1 = round(mean(Tsoil_lev1) - 273.1, 2),
              Tsoil_lev2 = round(mean(Tsoil_lev2) - 273.1, 2),
              VWC_lev1 = round(mean(VWC_lev1), 3),
              VWC_lev2 = round(mean(VWC_lev2), 3),
              Precip = round(sum(Precip) * 1000, 2),
              LAI_high = round(max(LAI_high), 2),
              .groups = "drop") ->
    cache_summary

srdb_sample %>% 
    left_join(cache_summary, by = c("Year", "Longitude", "Latitude")) ->
    srdb_sample

# A few sanity check plots
ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = Tair),
               na.rm = TRUE) + 
    ggtitle("SRDB with ERA5 2m air temperature")

ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = Precip),
               na.rm = TRUE) + 
    ggtitle("SRDB with ERA5 total precipitation")

ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = LAI_high),
               na.rm = TRUE) + 
    ggtitle("SRDB with ERA5 max high LAI")

```

## Soil type data

```{r read-soil-types}
#| warning: false
#| fig-width: 12

# Extract for SRDB points
x <- rast("data-ERA5/ERA5_soil_type.grib")
plot(x)
y <- terra::extract(x, coords) # not time-varying, so fast
names(y) <- c("ID", "Soil_type_number")

# Attach names to numeric codes
y$Soil_type_number <- as.character(y$Soil_type_number) # for join below
soil_names <- tibble(strings = strsplit(units(x), "; ")[[1]]) %>% 
    separate(strings, into = c("Soil_type_number", "Soil_type_name"), convert = TRUE, sep = "=")
y <- y %>% left_join(soil_names, by = "Soil_type_number")

srdb_sample <- bind_cols(srdb_sample, y[-1])
```

## MODIS annual NPP

```{r modis-npp}
#| fig-width: 12

library(MODISTools)

# Load cached data if available
MODIS_NPP_CACHE <- "caches/MODIS_NPP_cache.parquet"
if(file.exists(MODIS_NPP_CACHE)) {
    message("Reading cache file ", MODIS_NPP_CACHE)
    cache <- arrow::read_parquet(MODIS_NPP_CACHE)    
} else {
    message("No MODIS NPP data cache found")
    cache <- tibble(Year = numeric(0), 
                    Longitude = numeric(0), Latitude = numeric(0),
                    MODIS_NPP = numeric(0))
}

# srdb_yrlonlat (the unique combinations of location and time)
# was already calculated above
cache_hits <- 0
for(i in seq_len(nrow(srdb_yrlonlat))) {
    yr <- srdb_yrlonlat$Year[i]
    lon <- srdb_yrlonlat$Longitude[i]
    lat <- srdb_yrlonlat$Latitude[i]
    
    # Are data already in cache?
    cache %>% 
        filter(Year == yr, Longitude == lon, Latitude == lat) ->
        dat
    if(nrow(dat) > 0) {
        #message("\t", nrow(dat), " data rows are in cache")
        cache_hits <- cache_hits + 1
    } else {
        # Use MODISTools to download the needed data
        if(cache_hits > 0) { 
            message("(", last_cache_hits, " cache hits)")
            cache_hits <- 0
        }
        message(i, "/", nrow(srdb_yrlonlat), ": ", yr, " ", lon, " ", lat, appendLF = FALSE)
        message("\tNo data in cache; querying MODIS NPP products...")
        npp <- NULL
        try({
        npp <- mt_subset("MOD17A3HGF", 
                         band = "Npp_500m", 
                         lat = lat, lon = lon, 
                         start = paste0(yr, "-01-01"), end = paste0(yr, "-12-31"))
        })
        if(is.null(npp)) {
            npp_dat <- tibble(Year = yr, 
                              Longitude = lon, Latitude = lat, 
                              MODIS_NPP = NA_real_)
        } else {
            npp_dat <- tibble(Year = yr, 
                              Longitude = lon, Latitude = lat, 
                              MODIS_NPP = npp$value)
        }
        
        cache <- bind_rows(cache, npp_dat)
        arrow::write_parquet(cache, MODIS_NPP_CACHE)
    }
}

cache %>% 
   # 32767 is the 'fill value' for this data product -- no measurement
    # There are also some other values in the 32760s used for non-vegetated surfaces
    # see https://lpdaac.usgs.gov/documents/972/MOD17_User_Guide_V61.pdf
    mutate(MODIS_NPP = if_else(MODIS_NPP >= 32760, NA_real_, MODIS_NPP / 10.0)) ->
    cache_summary

# Join and make a sanity-check plot
srdb_sample %>% 
    left_join(cache_summary, by = c("Year", "Longitude", "Latitude")) ->
    srdb_sample

ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = MODIS_NPP),
               na.rm = TRUE) + 
    ggtitle("SRDB with MOD17A3HGF NPP")

```

## MODIS 8-day GPP

This is *also* slow.

```{r modis-gpp}
#| fig-width: 12

# Load cached data if available
MODIS_GPP_CACHE <- "caches/MODIS_GPP_cache.parquet"
if(file.exists(MODIS_GPP_CACHE)) {
    message("Reading cache file ", MODIS_GPP_CACHE)
    cache <- arrow::read_parquet(MODIS_GPP_CACHE)    
} else {
    message("No MODIS GPP data cache found")
    cache <- tibble(Year = numeric(0), 
                    calendar_date = character(0),
                    Longitude = numeric(0), Latitude = numeric(0),
                    MODIS_GPP = numeric(0))
}

# srdb_yrlonlat (the unique combinations of location and time)
# was already calculated above
cache_hits <- 0
for(i in seq_len(nrow(srdb_yrlonlat))) {
    yr <- srdb_yrlonlat$Year[i]
    lon <- srdb_yrlonlat$Longitude[i]
    lat <- srdb_yrlonlat$Latitude[i]
    
    # Are data already in cache?
    cache %>% 
        filter(Year == yr, Longitude == lon, Latitude == lat) ->
        dat
    if(nrow(dat) > 0) {
        #message("\t", nrow(dat), " data rows are in cache")
        cache_hits <- cache_hits + 1
    } else {
        # Use MODISTools to download the needed data
        if(cache_hits > 0) { 
            message("(", cache_hits, " cache hits)")
            cache_hits <- 0
        }
        
        message(i, "/", nrow(srdb_yrlonlat), ": ", yr, " ", lon, " ", lat, appendLF = FALSE)
        message("\tNo data in cache; querying MODIS GPP products...")
        gpp <- NULL
        try({
        gpp <- mt_subset("MOD17A2HGF", 
                         band = "Gpp_500m", 
                         lat = lat, lon = lon, 
                         start = paste0(yr, "-01-01"), end = paste0(yr, "-12-31"))
        })
        if(is.null(gpp)) {
            gpp_dat <- tibble(Year = yr, 
                              calendar_date = NA_character_,
                              Longitude = lon, Latitude = lat, 
                              MODIS_GPP = NA_real_)
        } else {
            gpp_dat <- tibble(Year = yr, 
                              calendar_date = gpp$calendar_date,
                              Longitude = lon, Latitude = lat,
                              MODIS_GPP = gpp$value)
        }
        
        cache <- bind_rows(cache, gpp_dat)
        arrow::write_parquet(cache, MODIS_GPP_CACHE)
    }
}

# Summarise cache, join, sanity check plot
cache %>% 
    # 32767 is the 'fill value' for this data product -- no measurement
    # There are also some other values in the 32760s used for non-vegetated surfaces
    # see https://lpdaac.usgs.gov/documents/972/MOD17_User_Guide_V61.pdf
    mutate(MODIS_GPP = if_else(MODIS_GPP >= 32760, NA_real_, MODIS_GPP)) %>% 
    group_by(Year, Longitude, Latitude) %>% 
    summarise(MODIS_GPP = sum(MODIS_GPP) / 10.0, .groups = "drop") ->
    cache_summary

srdb_sample %>% 
    left_join(cache_summary, by = c("Year", "Longitude", "Latitude")) ->
    srdb_sample

ggplot(world) + 
    geom_sf() + 
    geom_point(data = srdb_sample, 
               aes(x = Longitude, y = Latitude, color = MODIS_GPP),
               na.rm = TRUE) + 
    ggtitle("SRDB with MOD17A2HGF GPP (annual)")

```

## Final joined dataset

```{r final-dataset}
#| fig-height: 8
#| fig-width: 10
srdb_sample %>% 
    select(-Study_midyear, -ID, -Year1) ->
    srdb_sample

srdb_sample %>% 
    select(Year, Rs_annual, Rh_annual, Rs_growingseason) %>% 
    pivot_longer(-Year, values_transform = as.character, values_drop_na = TRUE) %>% 
    mutate(Type = "1. IVs") ->
    srdb_ivs

srdb_sample %>% 
    select(Year, Ecosystem_type, SPEI12, 
           Tair, 
           MODIS_NPP, MODIS_GPP) %>% 
    pivot_longer(-Year, values_transform = as.character, values_drop_na = TRUE) %>% 
    mutate(Type = "2. Predictors") %>% 
    bind_rows(srdb_ivs) ->
    srdb_predictors

ggplot(srdb_predictors, aes(Year, name)) + 
    geom_bin2d() + 
    facet_grid(Type ~., scales = "free_y", space = "free_y")

message("All observations: ", nrow(srdb_sample))
message("Pre-2001 observations: ", sum(srdb_sample$Year < 2001))
message("Pre-1985 observations: ", sum(srdb_sample$Year < 1985))

na_data <- is.na(srdb_sample$SPEI12)
if(sum(na_data) > 0) {
    message("Removing ", sum(na_data), " rows with no climate/SPEI data")
    srdb_sample <- srdb_sample[!na_data,]
}

OUTFILE <- "srdb_joined_data.csv"
message("Writing ", OUTFILE)
write_csv(srdb_sample, OUTFILE)

DT::datatable(srdb_sample)
```

## Reproducibility

```{r}
#| echo: false

sessionInfo()
```
